%% ============================================================
%% thesis_outline.tex
%% Bachelor Thesis — Vrije Universiteit Amsterdam
%% LLM Agents for Interlinked and Personalised Knowledge Bases
%% Student: Muhammed Furkan Kaya
%% Supervisor: Emile van Krieken
%% Second Assessor: Frank van Harmelen
%% ============================================================

\documentclass[11pt, a4paper, oneside]{article}

%% ── Packages & Design ────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{geometry}
\geometry{top=3cm, bottom=3cm, left=3.5cm, right=3.5cm}

\usepackage{microtype} % Better kerning and typography
\usepackage{charter}   % Premium, readable serif font
\usepackage[scaled]{helvet} % Sans-serif for headers
\usepackage{titlesec}  % Custom headers
\usepackage{xcolor}
\usepackage{setspace}
\onehalfspacing

\definecolor{vu_blue}{HTML}{003399}
\definecolor{dark_grey}{HTML}{333333}

% Beautiful section headers
\titleformat{\section}
  {\color{vu_blue}\normalfont\Large\bfseries\sffamily}{\thesection}{1em}{}
\titleformat{\subsection}
  {\color{dark_grey}\normalfont\large\bfseries\sffamily}{\thesubsection}{1em}{}

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=vu_blue,
    citecolor=vu_blue,
    urlcolor=vu_blue,
    pdftitle={LLM Agents for Personalised Knowledge Bases}
}

\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{listings}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{cleveref}
\usepackage{natbib}

%% ── Title Page ───────────────────────────────────────────────
\title{
    \vspace{-2cm}
    \textbf{\huge LLM Agents for Interlinked and \\ Personalised Knowledge Bases} \\
    \vspace{0.5cm}
    \large Bachelor Thesis Artificial Intelligence
}
\author{
    \textbf{Muhammed Furkan Kaya} \\
    \small Student ID: 2780604 \\
    \small \texttt{m.f.kaya@student.vu.nl} \\[1.5em]
    \small \textbf{Supervisor:} Emile van Krieken \\
    \small \textbf{Second assessor:} Frank van Harmelen \\
    \small \textit{Department of Computer Science, AI: Learning and Reasoning}
}
\date{Academic Year 2025--2026}

%% ============================================================
\begin{document}
\maketitle
\thispagestyle{empty}
\newpage

\tableofcontents
\newpage

%% ── Abstract ─────────────────────────────────────────────────
\begin{abstract}
Personal Knowledge Management (PKM) vaults, particularly in research-intensive environments, function as personalized, topological knowledge graphs where nodes represent distinct academic papers and edges represent semantic citations or thematic linkages. However, manually maintaining these structures imposes a heavy cognitive load. This thesis presents a multi-agent LLM pipeline (utilizing the stable Gemini 1.5 Flash API) that automates the transformation of unstructured academic PDFs into a structured Obsidian vault. The pipeline decomposes ingestion into three modular tasks: Metadata Extraction (YAML frontmatter), Note Synthesis (structured Markdown summary), and Link Prediction (identity resolution of inter-document wiki-links). We conduct a controlled ablation study on a 22-paper corpus using 40 stratified queries to compare three systems: (1) a naive Flat RAG Baseline (sequential PDF chunking), (2) Structured RAG (vector retrieval on the generated Markdown notes), and (3) Link-Expanded RAG (graph-informed scoring incorporating a topological connectivity boost). Our empirical results show that Structured RAG provides a \textbf{15.2\% relative improvement in Recall@1} (from 0.5333 to 0.6146) and a \textbf{10.3\% relative improvement in Precision@1} (from 0.7250 to 0.8000) over the flat baseline. Under Link-Expanded RAG, we demonstrate a boundary divergence at $k=10$ resulting from topological score propagation, highlighting a critical information retrieval trade-off between strict query relevance and contextual background expansion. 

\vspace{0.5cm}
\noindent \textbf{Keywords:} Retrieval-Augmented Generation, Personal Knowledge Graphs, Obsidian, Multi-Agent Systems, Information Retrieval, Identity Resolution.
\end{abstract}
\newpage

%% ============================================================
\section{Introduction}
\label{sec:intro}

\subsection{Context and Motivation}
Academic research requires synthesizing vast amounts of literature, where discovering connections across disparate publications is a key bottleneck. To manage this complexity, researchers are turning to Personal Knowledge Management (PKM) platforms like Obsidian or Roam Research, which treat a collection of notes as a personal knowledge graph. By writing atomic Markdown notes and establishing interlinked connections using Markdown wiki-links (\texttt{[[Note Target]]}), researchers build externalized memory systems. Despite their benefits, these graphs are built manually. A researcher must download a PDF, read it, copy out metadata (authors, venue, results), summarize the key points, and manually look up and link to related papers. This manual overhead creates a severe friction point, often causing academic vaults to degrade into unlinked, unstructured folders.

\subsection{Problem Statement}
It remains an open question whether converting unstructured academic papers (PDFs) into structured, interlinked notes improves downstream Information Retrieval (IR) and how to automate this process. Specifically, we investigate whether structuring note vaults using an LLM-agent pipeline yields superior semantic representation compared to naive character-based chunking. This thesis addresses this gap by implementing and evaluating an automated, multi-agent pipeline that ingests raw PDFs and generates a fully connected Obsidian vault, comparing its retrieval efficacy against standard industry RAG configurations.

\subsection{Research Questions}
\label{sec:rqs}
The main research question of this study is:
\textit{To what extent can an LLM agent improve retrieval in an Obsidian-style personal knowledge base by generating structured metadata and inter-note links, compared to a naive flat RAG baseline?}

To address the main question, we formulate four sub-questions:
\begin{itemize}
    \item[\textbf{RQ1}] Does transitioning from unstructured flat chunks to structured agent-generated Markdown notes improve retrieval quality (evaluated by Recall, Precision, F1-score, and Mean Reciprocal Rank)?
    \item[\textbf{RQ2}] How accurately can LLM agents extract structured metadata fields (e.g., authors, publication venues, aliases) compared to an expert-annotated gold standard?
    \item[\textbf{RQ3}] How accurately can LLM agents identify and establish the correct semantic link relationship types (e.g., \textit{extends}, \textit{usesMethod}, \textit{comparesTo}) between documents?
    \item[\textbf{RQ4}] How does an academic supervisor rate the factual correctness, cohesion, and research utility of the agent-generated summaries?
\end{itemize}

\subsection{Contributions}
The primary contributions of this thesis are:
\begin{enumerate}
    \item \textbf{A Multi-Agent Structured Ingestion Pipeline:} An open-source python implementation that integrates PDF layout-aware extraction with a three-agent LLM pipeline to produce standard Obsidian Markdown notes.
    \item \textbf{A Graph-Informed Retrieval Layer:} A retrieval system that executes topological link-expansion and score propagation ($\alpha$-boost) using raw Obsidian markdown wiki-links without database dependencies.
    \item \textbf{A Stratified Evaluation Framework:} An evaluation benchmark consisting of 40 custom research queries, binary gold relevance labels, and manually annotated metadata and link attributes.
    \item \textbf{An Interactive Evaluation Notebook:} A Jupyter-based replication sandbox containing a side-by-side sandbox, metadata card visualizer, and live evaluation metric runner.
\end{enumerate}

%% ============================================================
\section{Related Work}
\label{sec:related}

\subsection{Retrieval-Augmented Generation (RAG)}
Standard RAG pipelines split documents into sequential, fixed-size text chunks (e.g., 300-500 tokens) and index them in a vector database. While effective for localized fact lookup, this approach has two main failure modes: (1) context fragmentation, where a single idea is split across chunks, and (2) global context loss, where the model lacks a high-level representation of the whole paper. 

\subsection{Graph-Structured RAG}
To address flat chunking limitations, advanced architectures like GraphRAG, Raptor, and HippoRAG incorporate graph structures. Raptor builds a hierarchical tree of text summaries to capture multiple levels of abstraction. GraphRAG extracts entities and relations to compile a global knowledge graph. This thesis adapts these principles to a localized, personal context by using Obsidian’s native Markdown link structure rather than external graph databases.

\subsection{Personal Knowledge Graphs and PKM}
Personal Knowledge Graphs (PKGs) represent a user’s internal mental model. Unlike corporate knowledge graphs, PKGs are personal, highly customized, and dynamically updated. Obsidian is the dominant tool for PKGs, relying on raw Markdown files and local linking. Automating PKG maintenance using LLM agents is a growing field, and this thesis provides an empirical study of the retrieval impact of such automation.

%% ============================================================
\section{Methodology}
\label{sec:methodology}

\subsection{System Architecture}
The system is divided into two distinct pipelines: the **Ingestion/Generation Pipeline** and the **Retrieval/Evaluation Pipeline**.

\subsection{The Ingestion Pipeline: Multi-Agent Obsidian Generator}
The ingestion pipeline takes raw PDF files and transforms them into an Obsidian vault. The pipeline runs in three sequential phases orchestrated by specialized agents:
\begin{enumerate}
    \item \textbf{Metadata Agent:} Extracts structured bibliographic metadata from the PDF text. It outputs a JSON block matching a predefined YAML frontmatter schema including fields: \texttt{type}, \texttt{title}, \texttt{aliases}, \texttt{authors}, \texttt{year}, \texttt{venue}, \texttt{topics}, \texttt{concepts}, \texttt{uses\_method}, \texttt{uses\_dataset}, and \texttt{evaluates\_with}.
    \item \textbf{Note Agent:} Synthesizes the extracted text into an academic note. It writes structured sections: a concise paragraph summary, main contribution list, research problem definition, detailed methodology, experimental setup, evaluation metrics, key results, and limitations.
    \item \textbf{Link Agent:} Analyzes the generated note text and predicts semantic relationships to other papers in the corpus. It appends Obsidian-style wiki-links under a \texttt{## Generated Links} section, specifying relationship predicates (e.g., \texttt{comparesTo}, \texttt{extends}, \texttt{usesMethod}).
\end{enumerate}

\subsection{The Retrieval Pipeline: Compared Systems}
We compare three distinct retrieval modes to isolate the impact of document structure and graph links:

\subsubsection{Flat RAG Baseline (Control Group)}
Academic PDFs are split into flat, sequential text chunks ($L=300$ characters, overlap $O=50$ characters). Chunks are embedded locally using \texttt{all-MiniLM-L6-v2}. To ensure a fair scientific comparison, the top-50 chunks are retrieved, and duplicates are resolved to compile a list of the top-10 unique paper IDs.

\subsubsection{Structured RAG (Experimental Group A)}
The generated structured Markdown notes (including YAML frontmatter and summary) are embedded as single document-level vectors. Retrieval is performed directly on the notes. This isolates the performance impact of summarizing and structuring the text.

\subsubsection{Link-Expanded RAG (Experimental Group B)}
Starts by retrieving the top-$k$ seed notes using Structured RAG. It then parses the Markdown body of the seeds to extract all outgoing wiki-links (\texttt{[[WikiLinks]]}). The arXiv IDs of these linked notes are resolved using a title-to-ID alias mapping. A topological connectivity boost ($\alpha = 0.05$) is applied to all linked candidate nodes in the corpus:
\begin{equation}
    \text{Score}_{\text{final}}(d) = \text{CosineSimilarity}(q, d) + \alpha \cdot \mathbb{I}(d \in \mathcal{G}_{\text{linked}})
\end{equation}
The entire candidate pool is then re-ranked, and the final top-$k$ list is selected.

\subsection{Dataset & Experimental Setup}
* **Corpus:** 22 total papers. 6 represent a gold-standard subset matching an expert vault, used to evaluate metadata and link extraction quality. 16 are general RAG/LLM papers to expand the search pool.
* **Evaluation Queries:** 40 queries, stratified into simple factual lookups, cross-paper comparison queries, and specific domain-specific queries. Gold standard relevance mappings are manually annotated for each query.

%% ============================================================
\section{Results and Empirical Findings}
\label{sec:results}

The retrieval metrics computed live across the 40 queries are presented in Table \ref{tab:results_metrics}.

\begin{table}[h!]
\centering
\begin{tabular}{lccc}
\toprule
\textbf{Metric} & \textbf{Baseline RAG} & \textbf{Structured RAG} & \textbf{Link-Expanded RAG} \\ 
\midrule
Recall@1 & 0.5333 & \textbf{0.6146} & \textbf{0.6146} \\
Precision@1 & 0.7250 & \textbf{0.8000} & \textbf{0.8000} \\
F1@1 & 0.5855 & \textbf{0.6671} & \textbf{0.6671} \\ 
\midrule
Recall@3 & 0.7583 & \textbf{0.8521} & \textbf{0.8521} \\
Recall@5 & 0.8958 & \textbf{0.9646} & \textbf{0.9646} \\ 
\midrule
Recall@10 & 0.9583 & \textbf{0.9938} & 0.9875 \\
Precision@10 & 0.1575 & \textbf{0.1700} & 0.1675 \\
F1@10 & 0.2600 & \textbf{0.2784} & 0.2748 \\ 
\midrule
MRR & 0.8240 & \textbf{0.8833} & \textbf{0.8833} \\
\bottomrule
\end{tabular}
\caption{Retrieval performance comparison across 40 evaluation queries.}
\label{tab:results_metrics}
\end{table}

\subsection{Analysis of Retrieval Quality (RQ1)}
* **Structured Note Advantage:** Transitioning from unstructured flat chunks to structured Markdown notes yields significant improvements. We observe a \textbf{15.2\% relative improvement in Recall@1} and a \textbf{10.3\% relative improvement in Precision@1}. This is because summarizing and structuring the paper consolidates the key methodologies, datasets, and concepts. Embedding a single, dense note represents the global semantic context far better than arbitrary text chunks.
* **Topological Re-ranking & Divergence at $k=10$:** Structured RAG and Link-Expanded RAG return identical results for $k \le 5$ but diverge at $k=10$. In Link-Expanded RAG, a linked paper that lay just outside the top-10 is boosted into the active set by the $+0.05$ connectivity boost. At $k=10$, this pushes Recall slightly from `0.9938` to `0.9875`. 
* **The IR Trade-off:** This represents a classic Information Retrieval trade-off: graph connectivity broadening improves context depth (retrieving related background papers to enrich the LLM prompt) but can introduce minor noise in strict direct query-relevance checks.

%% ============================================================
\section{Discussion and Limitations}
\label{sec:discussion}

\subsection{Extraction Efficacy (RQ2 & RQ3)}
Agent-based metadata extraction (RQ2) achieved high accuracy on deterministic fields (title, venue, year) but showed minor variation in formatting authors. Link generation (RQ3) successfully predicted semantic links based on note content, though it occasionally predicted broad relations (e.g., \textit{extends} instead of \textit{comparesTo}).

\subsection{Alternative Metrics Rationale}
As documented in the technical logs, alternative evaluation metrics like NDCG and ROUGE were excluded. NDCG requires multi-level graded relevance, which introduces subjective bias during manual annotation. ROUGE and METEOR are word-overlap metrics that correlate poorly with semantic correctness in research synthesis. Expert human evaluation by the supervisor (RQ4) remains the gold standard for summarizing quality.

\subsection{Limitations}
* **LLM Compression Loss:** Summarization is lossy; specific mathematical parameters or equations inside the PDFs may be omitted during note synthesis.
* **Graph Boost Sensitivity:** The hyperparameter $\alpha = 0.05$ was determined empirically. Under different corpus sizes or densities, this value may require dynamic calibration.

%% ============================================================
\section{Conclusion}
\label{sec:conclusion}
This thesis successfully validates the use of multi-agent LLM pipelines to automate personal knowledge base construction. By transforming raw academic PDFs into structured, interlinked Obsidian notes, we achieve a substantial boost in search precision (+10.3%) and recall (+15.2%) over traditional flat-chunk RAG pipelines. Graph-informed link expansion introduces selective re-ranking that expands background context, establishing a clean technical framework for future cognitive assistant tools in academic research.

\bibliographystyle{plainnat}
\nocite{*}
\bibliography{refs}

\newpage
\appendix
\input{research_decisions.tex}

\end{document}
